\begin{abstract}
Smart contracts, primarily driven by blockchains, are increasingly adopted in practice to automate agreements, reduce costs, and improve efficiency.
Nevertheless, writing efficient and secure smart contracts still requires expertise and a deep understanding of low-level implementations. Declarative smart contracts, based on Datalog-style languages, aim to simplify deployment by allowing developers to focus on high-level contract logic. While existing tools translate declarative specifications into executable code, they still face limitations in expressiveness and efficiency. This paper presents \proto, a framework that extends declarative smart contract specifications and applies a set of optimization techniques to automatically generate gas-efficient contract implementations. \proto enhances expressiveness by supporting advanced expressions and on-chain context variables, enabling a broader range of contract types. To optimize gas consumption, \proto adapts classical query optimization techniques and proposes an enhanced deterministic selective view materialization algorithm over view elimination. A profiling-based simulation framework further ensures accurate but cost-free gas estimation without relying on static cost models. 
We build a declarative benchmark using top-deployed real-world contracts. Our experimental evaluation reveals that \proto supports diverse contract logic and significantly improves gas efficiency, especially by reducing storage overhead. This narrows the performance gap between auto-generated and expert-hand-crafted contracts, offering a practical path for adopting declarative programming in smart contract development.
\end{abstract}